---
title: "Geometry-consistent statistical workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Geometry-consistent statistical workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", error = FALSE,
                     fig.width = 5, fig.height = 4)
library(Riemann)
set.seed(1701)
```

A geometry specifies the distances and operations used by an analysis. Select it
before fitting, inspect the numerical result, and reuse the fitted object for new
observations. The examples below use small synthetic observations. They illustrate
software behavior, without making scientific claims about an application dataset.

## Validate observations and select a geometry

Construct positive-definite matrices by exponentiating real symmetric matrices.
The off-diagonal chart entries give noncommuting observations.

```{r spd-data}
matrix_exp <- function(a) {
  e <- eigen(a, symmetric = TRUE)
  e$vectors %*% diag(exp(e$values)) %*% t(e$vectors)
}
t <- seq(-0.6, 0.6, length.out = 12)
matrices <- lapply(t, function(a) {
  matrix_exp(matrix(c(a, 0.2 * sin(3 * a), 0.2 * sin(3 * a), -0.5 * a), 2))
})
x <- wrap.spd(matrices)
airm <- riem.geometry(x, "affine_invariant")
log_euclidean <- riem.geometry(x, "log_euclidean")
airm
log_euclidean
```

Affine-invariant and log-Euclidean distances describe different problems. The
legacy SPD names `"intrinsic"` and `"extrinsic"` resolve to these two choices,
respectively. The capability table records which operation routes are available;
its validation status also identifies families still awaiting a broader audit.

```{r capabilities}
capability <- riem.capabilities()
capability[capability$manifold_id %in% c("spd", "sphere", "landmark"), ]
```

Positive definiteness is checked independently of full rank. This deliberately
invalid example must produce an error; the vignette verifies that the error occurs.

```{r invalid-spd}
invalid <- tryCatch(wrap.spd(list(diag(c(-1, 1)))), error = identity)
stopifnot(inherits(invalid, "error"))
conditionMessage(invalid)
```

## Inspect a numerical summary

The shared solver records its objective and termination state. A finite matrix by
itself is not a convergence certificate.

```{r summaries}
m <- riem.mean(x, geometry = airm, maxiter = 200, eps = 1e-8, trace = TRUE)
m
stopifnot(m$converged)
c(objective = m$objective, iterations = m$iterations)
mlog <- riem.mean(x, geometry = log_euclidean)
stopifnot(mlog$converged)
riem.median(x, geometry = log_euclidean, maxiter = 200)
```

## Fit, predict, and reconstruct tangent components

Tangent PCA approximates variation near a trained reference. It retains a
metric-preserving coordinate convention and any coordinate offset. Its explained
variance is a property of this tangent representation, rather than exact global
variance on the manifold.

```{r spd-pca}
pca <- riem.pga(x, ndim = 2, geometry = log_euclidean)
pca
summary(pca)
stopifnot(isTRUE(all.equal(unname(predict(pca, x)), unname(pca$embed))))
new_x <- wrap.spd(list(matrix_exp(matrix(c(0.15, 0.08, 0.08, -0.1), 2))))
new_scores <- predict(pca, new_x)
new_scores
approximation <- riem.reconstruct(pca, new_scores)
approximation$data[[1]]
stopifnot(min(eigen(approximation$data[[1]], symmetric = TRUE)$values) > 0)
```

Prediction and reconstruction use the saved reference. With fewer components than
the data's rank, reconstruction is an approximation. Requests above numerical rank
are truncated with a warning and the effective dimension is recorded.

The sphere uses an orthonormal tangent basis. The following observations lie in a
small neighborhood where the reference and logarithms are well defined.

```{r sphere-pca}
sphere <- wrap.sphere(cbind(rep(1, 10), matrix(rnorm(20, sd = 0.08), 10, 2)))
sphere_pca <- riem.pga(sphere, ndim = 2, geometry = "round")
reconstructed <- riem.reconstruct(sphere_pca)
stopifnot(all(vapply(reconstructed$data, function(a) abs(sum(a^2) - 1) < 1e-10,
                    logical(1))))
plot(sphere_pca)
```

Antipodal logarithms are nonunique. Unsupported geometry-operation combinations
also produce explicit errors rather than silently switching metrics.

```{r unsupported-tangent}
unsupported <- tryCatch(riem.pga(sphere, geometry = "chordal"), error = identity)
stopifnot(inherits(unsupported, "error"))
conditionMessage(unsupported)
```

## Reuse a clustering model

The default Lloyd driver uses the shared mean solver and records each start. Labels
returned by prediction refer to the final trained centers.

```{r clustering}
set.seed(1702)
clusters <- riem.kmeans(x, k = 2, geometry = log_euclidean, nstart = 2)
clusters
clusters$starts
stopifnot(identical(predict(clusters, x), clusters$cluster))
predict(clusters, new_x)
```

## Tune scalar-response regression using every held-out fold

Here the predictor is a matrix and the response is numeric. Geometry is fixed
before tuning. The fold assignments are stored and every observation contributes
once to the held-out sum of squared errors. For grouped or temporal data, supply
splits appropriate to the sampling design. If preprocessing is learned from data,
it must be repeated within the corresponding training folds.

```{r regression}
y <- sin(2 * t) + rnorm(length(t), sd = 0.03)
regression <- riem.m2skregCV(
  x, y, geometry = log_euclidean, bandwidths = c(0.1, 0.25, 0.5),
  foldid = rep(1:3, length.out = length(t))
)
regression$errors
regression$fold_errors
prediction <- predict(regression, new_x, diagnostics = TRUE)
prediction
stopifnot(identical(regression$geometry$geometry_id, "log_euclidean"))
```

A finite prediction can still have weak local support. Effective neighborhood size
and distance relative to bandwidth help assess that limitation; numerical
stability does not establish statistical support.

## Preserve the fitted state

```{r persistence}
path <- tempfile(fileext = ".rds")
saveRDS(pca, path)
restored <- readRDS(path)
unlink(path)
stopifnot(isTRUE(all.equal(predict(restored, new_x), new_scores)))
sessionInfo()
```

Regular landmark tangent workflows use equivalence under the full orthogonal
group, including reflections, and align new observations to the trained reference.
Singular alignments and reconstructions outside the supported local chart are
rejected. This convention must agree with the scientific meaning of the landmarks;
it is not an orientation-preserving shape convention.
